# Neural Network with MXNet in Five Minutes

This is the first tutorial for new users of the R package `mxnet`. You will learn to construct a neural network to do regression in 5 minutes.

We will show you how to do classification and regression tasks respectively. The data we use comes from the package `mlbench`.

## Classification

First of all, let us load in the data and preprocess it:

```{r}
require(mlbench)
require(mxnet)

data(Sonar, package = "mlbench")

Sonar[,61] <- as.numeric(Sonar[,61])-1
train.ind <- c(1:50, 100:150)
train.x <- data.matrix(Sonar[train.ind, 1:60])
train.y <- Sonar[train.ind, 61]
test.x <- data.matrix(Sonar[-train.ind, 1:60])
test.y <- Sonar[-train.ind, 61]
```

Next we are going to use a multi-layer perceptron (MLP) as our classifier.
In `mxnet`, we have a function called `mx.mlp` so that users can build a general multi-layer neural network to do classification (`out_activation="softmax"`) or regression (`out_activation="rmse"`).
Note for the `softmax` activation, the output is zero-indexed not one-indexed. In the data we use:

```{r}
table(train.y)
table(test.y)
```

There are several parameters we have to feed to `mx.mlp`:

- Training data and label.
- Number of hidden nodes in each hidden layers.
- Number of nodes in the output layer.
- Type of the activation.
- Type of the output loss.
- The device to train `mx.gpu()` for GPU or `mx.cpu()` for CPU.
- Other parameters for `mx.model.FeedForward.create`.

The following code piece is showing a possible usage of `mx.mlp`:

```{r}
mx.set.seed(0)
model <- mx.mlp(train.x, train.y, hidden_node=10, out_node=2, out_activation="softmax",
                num.round=20, array.batch.size=15, learning.rate=0.07, momentum=0.9, 
                eval.metric=mx.metric.accuracy)
```

Note that `mx.set.seed` is the correct function to control the random process in `mxnet`. You can see the accuracy in each round during training. It is also easy to make prediction and evaluate.

To get an idea of what is happening, we can easily view the computation graph from R.

```{r}
graph.viz(model$symbol)
```

```{r}
preds <- predict(model, test.x)
pred.label <- max.col(t(preds)) - 1
table(pred.label, test.y)
```

Note for multi-class prediction, mxnet outputs `nclass` x `nexamples`, each each row corresponding to probability of that class.




<!-- INSERT SOURCE DOWNLOAD BUTTONS -->
